---
title: "Bayesian Concept Learning"
author: "omikad"
date: "April 4, 2016"
output: html_document
---

```{r echo=FALSE, results='hide'}
library(ggplot2)
library(knitr)
opts_chunk$set(echo=FALSE)
```

Hypothesis data consists of the following colummns:

- hypothesis: name of hypothesys

- prior.score: subjective value for hypotheseses. Lower value for "conceptually unnatural" prior

- prior: weighted prior.score

- expression: R expression to generate hypothesis values

- size: number of elements in hypothesis set

```{r}
hyps <- read.csv("hypothesis.csv", sep=";")

hyps$expression <- as.character(hyps$expression)

hyps$prior <- hyps$prior.score / sum(hyps$prior.score)

hyps$hypothesis <- factor(hyps$hypothesis, levels=rev(hyps$hypothesis))

hyps.set <- list()
for (i in 1:nrow(hyps)) {
  items <- eval(parse(text=hyps$expression[i]))
  hyps.set[[i]] <- items
  hyps$size[i] = length(items)
}

isInHypothesis <- function(x) {
  unlist(lapply(hyps.set, function(s){ all(x %in% s) }))
}

print(hyps)
```

Priors:

```{r}
ggplot(hyps, aes(x=hypothesis, y=prior)) + 
  ylim(0, 0.2) + 
  geom_bar(stat="identity") + 
  coord_flip() + 
  theme(panel.background=element_blank())
```

After observation of {16} we get likelihood:

```{r}
hyps$likelihood <- 1 / hyps$size * isInHypothesis(16)

ggplot(hyps, aes(x=hypothesis, y=likelihood)) + 
  ylim(0, 0.4) + 
  geom_bar(stat="identity") + 
  coord_flip() + 
  theme(panel.background=element_blank())
```

And posterior:

```{r}
hyps$posterior <- hyps$prior * hyps$likelihood
hyps$posterior <- hyps$posterior / sum(hyps$posterior)

ggplot(hyps, aes(x=hypothesis, y=posterior)) + 
  ylim(0, 0.4) + 
  geom_bar(stat="identity") + 
  coord_flip() + 
  theme(panel.background=element_blank())
```


After observation of {16,8,2,64} we get likelihood:

```{r}
d <- c(16,8,2,64)
hyps$likelihood <- 1 / (hyps$size ^ length(d)) * isInHypothesis(d)

ggplot(hyps, aes(x=hypothesis, y=likelihood)) +  
  geom_bar(stat="identity") + 
  coord_flip() + 
  theme(panel.background=element_blank())
```

And posterior:

```{r}
hyps$posterior <- hyps$prior * hyps$likelihood
hyps$posterior <- hyps$posterior / sum(hyps$posterior)

ggplot(hyps, aes(x=hypothesis, y=posterior)) + 
  geom_bar(stat="identity") + 
  coord_flip() + 
  theme(panel.background=element_blank())
```